---
title: "Case Study Report: From Linear Regression to Geostatistical Regression"
subtitle: "Fundamentals of Spatial Data Engineering (FsDE)"
author: "Student Name (Student ID)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
bibliography: references.bib
link-citations: true
---

# 1. Introduction

*Background & problem.* Briefly describe the study area and motivate why modeling rainfall variability matters for geospatial analysis (e.g., interpolation, hydrology, climate). Clearly state the **geographical/geospatial questions** your report answers, such as:

-   *Q1.* How well do elevation (DEM) and NDVI individually explain rainfall variability?
-   *Q2.* Does a multiple regression (DEM + NDVI) improve explanatory power vs. single-predictor models?
-   *Q3.* Does accounting for spatial autocorrelation via **GLS** improve predictive error vs **OLS**?
-   *Q4.* Which geostatistical predictor is better under LOOCV: **Ordinary Kriging (OK)** or **Regression Kriging (RK)**?

*Report outline.* Close with a short paragraph describing the structure: Methods (data, workflow), Results (per question), Discussion, and Conclusions. Cite core references for kriging and geostatistics as needed \cite{cressie1993,hengl2009,pebesma2004}.

# 2. Methodology

## 2.1 Workflow and spatial analysis

Briefly summarize the steps; include (or later paste) a flowchart image showing:

1)  Data ingestion & preprocessing → 2) Covariate extraction (DEM, NDVI) → 3) OLS/Multiple OLS → 4) Variogram fitting (residuals & rainfall) → 5) GLS → 6) LOOCV for OK & RK → 7) Metrics & comparison.

> *Tip:* You can create and export a flowchart figure separately (e.g., draw.io, DiagrammeR) and include it with `knitr::include_graphics()`.

## 2.2 Data Management Plan

Describe the datasets (station rainfall, Swiss boundary, DEM, NDVI): CRS, resolution, units, temporal coverage, licensing/source, preprocessing. Mention file organization and reproducibility practices.

## 2.3 Methods

### 2.3.1 Packages

```{r setup, message=FALSE, warning=FALSE}
# Install once if needed:
# install.packages(c("sf","raster","sp","spdep","gstat","gridExtra","cowplot"))

library(sf)
library(raster)
library(sp)
library(spdep)
library(gstat)
library(gridExtra)
library(cowplot)

# For reproducibility in any random ops
set.seed(42)
```

### 2.3.2 Load and organize data

```{r data, message=FALSE, warning=FALSE}
# Adjust to your local path:
setwd("C:/Users/darey/OneDrive/Documents/MGEO/MGEO/FromLinearRegressionToGeostatisticalRegression/Data")

rain.DF <- read.table("sic_obs.dat", sep = ",", header = TRUE)

Boundary.sf <- st_read("Swiss_Bound.shp")
Boundary.sp <- as(Boundary.sf, "Spatial")

DEM  <- raster("DEM.grd")
NDVI <- raster("NDVI.grd")

# Promote table to SpatialPointsDataFrame
rain.SpDF <- rain.DF
coordinates(rain.SpDF) <- c("x_coor","y_coor")

# Extract covariates at station locations
x1 <- extract(DEM,  rain.SpDF, df = TRUE)[,2]  # Elevation
x2 <- extract(NDVI, rain.SpDF, df = TRUE)[,2]  # NDVI
y  <- rain.SpDF@data$rain_mm                   # Response

# Design matrix (intercept, DEM, NDVI)
X <- cbind(1, x1, x2)

# Store back into attributes
rain.SpDF@data$y  <- y
rain.SpDF@data$x1 <- x1
rain.SpDF@data$x2 <- x2
```

### 2.3.3 Task 1 — Simple OLS models (`y ~ x1` vs `y ~ x2`)

```{r task1-fit}
LM_x1 <- lm(y ~ x1)
LM_x2 <- lm(y ~ x2)

rmse <- function(m) sqrt(mean(residuals(m)^2))

ols_simple_metrics <- data.frame(
  Model   = c("y ~ x1 (DEM)", "y ~ x2 (NDVI)"),
  SSE     = c(sum(residuals(LM_x1)^2), sum(residuals(LM_x2)^2)),
  MSE     = c(mean(residuals(LM_x1)^2), mean(residuals(LM_x2)^2)),
  RMSE    = c(rmse(LM_x1), rmse(LM_x2)),
  R2      = c(summary(LM_x1)$r.squared, summary(LM_x2)$r.squared),
  Adj_R2  = c(summary(LM_x1)$adj.r.squared, summary(LM_x2)$adj.r.squared)
)
ols_simple_metrics
```

### 2.3.4 Task 2 — Multiple OLS (`y ~ x1 + x2`) and comparison

```{r task2-fit}
LM_both <- lm(y ~ x1 + x2)

ols_all_metrics <- data.frame(
  Model   = c("y ~ x1 (DEM)", "y ~ x2 (NDVI)", "y ~ x1 + x2"),
  SSE     = c(sum(residuals(LM_x1)^2),
              sum(residuals(LM_x2)^2),
              sum(residuals(LM_both)^2)),
  MSE     = c(mean(residuals(LM_x1)^2),
              mean(residuals(LM_x2)^2),
              mean(residuals(LM_both)^2)),
  RMSE    = c(rmse(LM_x1), rmse(LM_x2), rmse(LM_both)),
  R2      = c(summary(LM_x1)$r.squared,
              summary(LM_x2)$r.squared,
              summary(LM_both)$r.squared),
  Adj_R2  = c(summary(LM_x1)$adj.r.squared,
              summary(LM_x2)$adj.r.squared,
              summary(LM_both)$adj.r.squared)
)
ols_all_metrics
```

### 2.3.5 Task 3 — OLS vs GLS (RMSE comparison)

```{r task3-gls, message=FALSE, warning=FALSE}
# 1) Empirical variogram of OLS residuals
e_ols   <- residuals(LM_both)
vgm.Emp.e <- variogram(e_ols ~ 1, locations = rain.SpDF)
vgmSph.e  <- fit.variogram(vgm.Emp.e, model = vgm(psill = 8000, model = "Sph", range = 60000, nugget = 500))

# 2) Build covariance Σ from residual variogram and perform GLS
DMat       <- spDists(rain.SpDF)
Sigma_ols1 <- variogramLine(vgmSph.e, dist_vector = DMat, covariance = TRUE)
Beta_gls1  <- solve(t(X) %*% solve(Sigma_ols1) %*% X) %*% (t(X) %*% solve(Sigma_ols1) %*% y)
e_gls1     <- as.vector(y - (X %*% Beta_gls1))

# 3) Iterate GLS to refine variogram and coefficients
K_iter <- 50
B.gls.ite <- matrix(NA_real_, nrow = K_iter, ncol = ncol(X))
for (i in 1:K_iter) {
  v_emp <- variogram(e_gls1 ~ 1, locations = rain.SpDF, covariogram = FALSE)
  v_fit <- fit.variogram(v_emp, model = vgm(8000, "Sph", 60000, nugget = 500))
  Sigma <- variogramLine(v_fit, dist_vector = DMat, covariance = TRUE)
  B.gls.ite[i, ] <- solve(t(X) %*% solve(Sigma) %*% X) %*% (t(X) %*% solve(Sigma) %*% y)
  e_gls1 <- as.vector(y - (X %*% B.gls.ite[i, ]))
}

# 4) Final GLS residuals
B.gls <- B.gls.ite[K_iter, ]
e.gls <- as.vector(y - (X %*% B.gls))

# 5) RMSE with residual df (k = 2 predictors)
n <- length(y); k <- 2
SSE_ols <- t(e_ols) %*% e_ols
MSE_ols <- SSE_ols / (n - (k + 1))
RMSE_ols <- sqrt(MSE_ols)

SSE_gls <- t(e.gls) %*% e.gls
MSE_gls <- SSE_gls / (n - (k + 1))
RMSE_gls <- sqrt(MSE_gls)

gls_cmp <- data.frame(
  Model = c("OLS (y ~ x1 + x2)", "GLS (y ~ x1 + x2)"),
  RMSE  = c(RMSE_ols, RMSE_gls)
)
gls_cmp
```

### 2.3.6 Task 4 — LOOCV: Ordinary Kriging vs Regression Kriging

```{r task4-cv, message=FALSE, warning=FALSE}
# Variogram for rainfall (OK & RK)
vgm.Emp.y <- variogram(y ~ 1, locations = rain.SpDF)
vgmExp.y  <- fit.variogram(vgm.Emp.y, model = vgm(psill = 15000, model = "Exp",
                                                  range = 60000, nugget = 500))

# LOOCV
OK.cv <- krige.cv(y ~ 1,           locations = rain.SpDF, model = vgmExp.y)
RK.cv <- krige.cv(y ~ x1 + x2,     locations = rain.SpDF, model = vgmExp.y)

cv_metrics <- function(cv_obj, y_obs = rain.SpDF$y) {
  res <- cv_obj$residual
  ME   <- mean(res)
  MSE  <- mean(res^2)
  RMSE <- sqrt(MSE)
  R2   <- 1 - var(res) / var(y_obs)
  data.frame(ME=ME, MSE=MSE, RMSE=RMSE, R2=R2)
}

OK_metrics <- cv_metrics(OK.cv)
RK_metrics <- cv_metrics(RK.cv)

cv_cmp <- rbind(
  cbind(Model="Ordinary Kriging", OK_metrics),
  cbind(Model="Regression Kriging", RK_metrics)
)
cv_cmp
```

# 3. Results

## 3.1 Task 1 — Comparing simple linear models

Summarize coefficients, significance, and the metrics table `ols_simple_metrics`. Explain which covariate better explains rainfall and why (e.g., higher R², lower RMSE).

## 3.2 Task 2 — Multiple OLS vs single-predictor models

Discuss `ols_all_metrics`. Note whether adding elevation to NDVI brings meaningful gains and how large they are (ΔRMSE, ΔR²).

## 3.3 Task 3 — OLS vs GLS

Interpret the `gls_cmp` RMSE comparison. When might GLS help, and why might it not here?

## 3.4 Task 4 — LOOCV: OK vs RK

Interpret `cv_cmp` (ME bias, RMSE, and R²). Explain conditions under which RK is expected to outperform OK.

# 4. Discussion

Reflect on validity and limitations: data density, station representativeness, CRS and support/resolution issues, model misspecification, variogram sensitivity, and uncertainty communication. Relate to broader frameworks or SDGs if relevant.

# 5. Conclusions

Provide concise answers to Q1–Q4. State key takeaways (best-performing approach under the given data) and what could improve results (denser data, additional covariates, nonstationary models, etc.).

# References

*Rendered automatically from* `references.bib` *based on in-text citations.*

# Appendix A. Detailed Data Descriptions

Add extended metadata and any additional checks not essential to the main text.

# Appendix B. Additional Figures or Maps

Include A4 (or larger) cartographic outputs following cartographic rules.
