---
title: "From Linear Regression to Geostatistical Regression — Reproducible R Tutorial"
author: "Your Name"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

> **About this document**\
> This R Markdown walks you from classical linear regression to geostatistical regression and kriging with the Swiss rainfall example. It mirrors the structure of the provided exercise and adds step‑by‑step explanations and code chunks you can run as‑is once your data files are in place.

## 0) Setup

### 0.1 Packages

``` {r}
# Core spatial + modeling stack
required_pkgs <- c(
  "sf", "sp", "raster", "gstat", "spdep",
  "gridExtra", "cowplot"
)

to_install <- setdiff(required_pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

invisible(lapply(required_pkgs, library, character.only = TRUE))
```

### 0.2 Data: expected files and working directory

Place the following files in a working folder and set it below:

-   `sic_obs.dat` — point observations with columns including `x_coor`, `y_coor`, and rainfall (e.g., `rain_mm`).
-   `Swiss_Bound.shp` (+ its `.dbf`, `.shx`, `.prj`) — study area boundary.
-   `DEM.grd` — elevation raster.
-   `NDVI.grd` — NDVI raster.

``` {r}
# Adjust this path to your local folder:
wd <- "~/Documents/SwissRainfallTutorial"
if (!dir.exists(wd)) dir.create(wd, recursive = TRUE)
setwd(wd)

# For reproducibility in small random operations (e.g., sampling grids):
set.seed(42)
```

### 0.3 Load and organize data

``` {r}
# 1) Read table of station observations
rain.DF <- read.table("sic_obs.dat", sep = ",", header = TRUE)

# 2) Promote to SpatialPointsDataFrame
rain.SpDF <- rain.DF
coordinates(rain.SpDF) <- c("x_coor","y_coor")
proj4string(rain.SpDF) <- CRS("+init=epsg:2056") # CH1903+/LV95 as an example; change to match data

# 3) Read boundary as sf and convert to Spatial
Boundary.sf <- st_read("Swiss_Bound.shp", quiet = TRUE)
Boundary.sp <- as(Boundary.sf, "Spatial")

# 4) Read rasters
DEM  <- raster("DEM.grd")
NDVI <- raster("NDVI.grd")

# 5) Extract covariates at station locations
x1 <- extract(DEM,  rain.SpDF, df = TRUE)[,2]  # elevation
x2 <- extract(NDVI, rain.SpDF, df = TRUE)[,2]  # NDVI

# 6) Response and design matrix
y <- rain.SpDF@data$rain_mm
X <- cbind(Intercept = 1, x1 = x1, x2 = x2)

# 7) Attach for convenience
rain.SpDF@data$y  <- y
rain.SpDF@data$x1 <- x1
rain.SpDF@data$x2 <- x2
```

## Part 1 — Classical Linear Regression

### 1.1 Variance–covariance matrix of (y, x1, x2)

We center each variable and compute $\Sigma = Z^\* (Z^\*)' / (n-1)$.

``` {r}
n       <- length(y)
y.star  <- y  - mean(y)
x1.star <- x1 - mean(x1)
x2.star <- x2 - mean(x2)

Z <- cbind(y.star, x1.star, x2.star)
Sigma.Z <- (t(Z) %*% Z) / (n - 1)
Sigma.Z
```

*Interpretation.* Diagonal entries are variances of each variable; off‑diagonals are covariances. Signs indicate direction of linear association.

### 1.2 Correlation matrix

``` {r}
CorMat <- matrix(NA_real_, nrow = 3, ncol = 3)
for (i in 1:3) for (j in 1:3) {
  CorMat[i,j] <- Sigma.Z[i,j] / sqrt(Sigma.Z[i,i] * Sigma.Z[j,j])
}
colnames(CorMat) <- rownames(CorMat) <- c("y","x1","x2")
CorMat
```

### 1.3 Least‑squares regression via `lm()` and matrix algebra

Model: $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \varepsilon_i$.

``` {r}
# Using lm()
LM1 <- lm(y ~ x1 + x2)
summary(LM1)

# Matrix estimator: beta_hat = (X'X)^(-1) X'y
Beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
as.vector(Beta_hat)
```

The `summary()` reports estimates, standard errors, t‑values, p‑values, residual standard error, R² and adjusted R².

### 1.4 Residual diagnostics

We check normality and homoscedasticity with histogram, Q–Q plot, and residuals vs fitted/observed/covariates.

``` {r}
par(mfrow = c(2,1))
e <- resid(LM1)
hist(e, nclass = 20, freq = FALSE, xlab = "Residuals", main = "Histogram of residuals")
abline(v = 0, lwd = 3, lty = 2, col = "orange")
qqnorm(e, main = "Q-Q plot of residuals"); qqline(e, lwd = 2, lty = 2, col = "orange")

par(mfrow = c(2,2))
y_cap <- fitted(LM1)
plot(y_cap, e, ylab = "Residuals", xlab = "Fitted values", main = "Residuals vs Fitted"); abline(h = 0, lwd = 2, lty = 2, col = "orange")
plot(y,     e, ylab = "Residuals", xlab = "Observed values", main = "Residuals vs Observed"); abline(h = 0, lwd = 2, lty = 2, col = "orange")
plot(x1,    e, ylab = "Residuals", xlab = "DEM (x1)", main = "Residuals vs x1"); abline(h = 0, lwd = 2, lty = 2, col = "orange")
plot(x2,    e, ylab = "Residuals", xlab = "NDVI (x2)", main = "Residuals vs x2"); abline(h = 0, lwd = 2, lty = 2, col = "orange")
par(mfrow = c(1,1))
```

### 1.5 SSE, MSE, RMSE

``` {r}
SSE <- sum((y - y_cap)^2)
K   <- 2  # number of predictors (x1, x2)
MSE <- SSE / (n - (K + 1))
RMSE <- sqrt(MSE)
SSE; MSE; RMSE
```

### 1.6 Covariance & correlation of coefficient estimates

Variance–covariance: $\widehat{\Sigma}_\beta = \widehat{\sigma}^2 (X'X)^{-1}$.

``` {r}
CovBeta  <- as.vector(MSE) * solve(t(X) %*% X)
CovBeta_lm <- vcov(LM1)           # should match
round(CovBeta, 6); round(CovBeta_lm, 6)

# Correlation between coefficients
p <- ncol(X)
CrrBeta <- matrix(NA_real_, nrow = p, ncol = p)
for (i in 1:p) for (j in 1:p) CrrBeta[i,j] <- CovBeta[i,j] / sqrt(CovBeta[i,i] * CovBeta[j,j])
colnames(CrrBeta) <- rownames(CrrBeta) <- colnames(X)
CrrBeta
```

### 1.7 Spatial prediction with the linear model (deterministic trend)

We predict across a grid using `predict.lm()` and visualize.

``` {r}
# Create a regular prediction grid within boundary
PredGridPts <- spsample(Boundary.sp, n = 200000, type = "regular")
gridded(PredGridPts) <- TRUE

# Extract covariates on the grid
x10 <- extract(DEM,  PredGridPts, df = TRUE)[,2]
x20 <- extract(NDVI, PredGridPts, df = TRUE)[,2]

newdata <- data.frame(x1 = x10, x2 = x20)
y0 <- predict(LM1, newdata = newdata, se.fit = TRUE, interval = "prediction")

# Wrap into SpatialPixelsDataFrame
y0.Sp <- SpatialPixelsDataFrame(points = PredGridPts, data = data.frame(fit = y0$fit[,"fit"], se = y0$se.fit))

# Plot
spplot(y0.Sp, zcol = "fit", main = "Linear model: spatial prediction (trend)")
spplot(y0.Sp, zcol = "se",  main = "Linear model: prediction standard error")
```

*Note.* This is a **deterministic** trend surface; it does not model spatial autocorrelation of the residuals.

## Part 2 — Geostatistical Exploration & Regression

### 2.1 Semivariograms of y and OLS residuals

We explore spatial autocorrelation with empirical semivariograms and fit legal models (Exp/Gau/Sph).

``` {r}
# Empirical semivariograms
vgm.cloud.y <- variogram(y ~ 1, locations = rain.SpDF, cloud = TRUE)
vgm.Emp.y   <- variogram(y ~ 1, locations = rain.SpDF)

vgm.cloud.e <- variogram(e ~ 1, locations = rain.SpDF, cloud = TRUE)
vgm.Emp.e   <- variogram(e ~ 1, locations = rain.SpDF)

grid.arrange(
  plot(vgm.cloud.y, main = "Semivariogram cloud — rainfall"),
  plot(vgm.Emp.y,   main = "Empirical semivariogram — rainfall"),
  plot(vgm.cloud.e, main = "Semivariogram cloud — OLS residuals"),
  plot(vgm.Emp.e,   main = "Empirical semivariogram — OLS residuals"),
  nrow = 2
)

# Fit models to rainfall semivariogram
vgmExp.y <- fit.variogram(vgm.Emp.y, model = vgm(psill = 15000, model = "Exp", range = 60000, nugget = 500))
vgmGau.y <- fit.variogram(vgm.Emp.y, model = vgm(psill = 15000, model = "Gau", range = 60000, nugget = 500))
vgmSph.y <- fit.variogram(vgm.Emp.y, model = vgm(psill = 15000, model = "Sph", range = 60000, nugget = 500))

grid.arrange(
  plot(vgm.Emp.y, vgmExp.y, main = "Rainfall — Exponential model"),
  plot(vgm.Emp.y, vgmGau.y, main = "Rainfall — Gaussian model"),
  plot(vgm.Emp.y, vgmSph.y, main = "Rainfall — Spherical model"),
  nrow = 2
)

# Fit models to residual semivariogram
vgmExp.e <- fit.variogram(vgm.Emp.e, model = vgm(psill = 8000,  model = "Exp", range = 60000, nugget = 500))
vgmGau.e <- fit.variogram(vgm.Emp.e, model = vgm(psill = 8000,  model = "Gau", range = 60000, nugget = 500))
vgmSph.e <- fit.variogram(vgm.Emp.e, model = vgm(psill = 8000,  model = "Sph", range = 60000, nugget = 500))

grid.arrange(
  plot(vgm.Emp.e, vgmExp.e, main = "Residuals — Exponential model"),
  plot(vgm.Emp.e, vgmGau.e, main = "Residuals — Gaussian model"),
  plot(vgm.Emp.e, vgmSph.e, main = "Residuals — Spherical model"),
  nrow = 2
)
```

*Reading the models.* Range \~ spatial correlation extent; sill \~ variance contribution; nugget \~ microscale variance/measurement error.

### 2.2 Geostatistical regression (GLS)

We iteratively estimate GLS coefficients using a covariance implied by the residual semivariogram.

``` {r}
# Distance matrix between observation locations
DMat <- spDists(rain.SpDF)

# Start from residual semivariogram model (e.g., spherical) to build covariance
Sigma.ols1 <- variogramLine(vgmSph.e, dist_vector = DMat, covariance = TRUE)

# First GLS coefficients
Beta.gls1 <- solve(t(X) %*% solve(Sigma.ols1) %*% X) %*% (t(X) %*% solve(Sigma.ols1) %*% y)

# Iterate: update residuals -> refit variogram -> recompute covariance -> update beta
B.gls.ite <- matrix(NA_real_, nrow = 100, ncol = ncol(X))
colnames(B.gls.ite) <- colnames(X)
e.gls1 <- as.numeric(y - (X %*% Beta.gls1))

for (i in 1:nrow(B.gls.ite)) {
  vgm.Emp.gls1 <- variogram(e.gls1 ~ 1, locations = rain.SpDF, covariogram = FALSE)
  vgmSph.gls1  <- fit.variogram(vgm.Emp.gls1, model = vgm(psill = 8000, model = "Sph", range = 60000, nugget = 500))
  Sigma.gls1   <- variogramLine(vgmSph.gls1, dist_vector = DMat, covariance = TRUE)
  B.gls.ite[i,] <- solve(t(X) %*% solve(Sigma.gls1) %*% X) %*% (t(X) %*% solve(Sigma.gls1) %*% y)
  e.gls1 <- as.numeric(y - (X %*% B.gls.ite[i,]))
}

# Inspect iteration paths
par(mfrow = c(2,2))
ts.plot(B.gls.ite[,1], lwd = 3, main = "GLS Intercept", xlab = "Iteration", ylab = "Estimate")
abline(h = tail(B.gls.ite[,1], 1), lwd = 2, lty = 2, col = "red")
ts.plot(B.gls.ite[,2], lwd = 3, main = "GLS x1", xlab = "Iteration", ylab = "Estimate")
abline(h = tail(B.gls.ite[,2], 1), lwd = 2, lty = 2, col = "red")
ts.plot(B.gls.ite[,3], lwd = 3, main = "GLS x2", xlab = "Iteration", ylab = "Estimate")
abline(h = tail(B.gls.ite[,3], 1), lwd = 2, lty = 2, col = "red")
par(mfrow = c(1,1))

# Final GLS estimates
B.gls <- matrix(tail(B.gls.ite, 1), nrow = ncol(X))
colnames(B.gls) <- "GLS"
B.gls
```

#### 2.2.1 Standard errors of GLS coefficients

``` {r}
# GLS residuals and error variance
e.gls   <- as.numeric(y - (X %*% B.gls))
SSE.gls <- sum(e.gls^2)
k <- 2
MSE.gls <- SSE.gls / (n - (k + 1))
RMSE.gls <- sqrt(MSE.gls)

# Covariance of GLS coefficients: (X' Σ^{-1} X)^{-1}
vgm.Emp.gls <- variogram(e.gls ~ 1, locations = rain.SpDF, covariogram = FALSE)
vgmSph.gls  <- fit.variogram(vgm.Emp.gls, model = vgm(psill = 8000, model = "Sph", range = 60000, nugget = 500))
Sigma.gls   <- variogramLine(vgmSph.gls, dist_vector = DMat, covariance = TRUE)

Sigma.Beta.Gls <- solve(t(X) %*% solve(Sigma.gls) %*% X)
SE.GLS <- sqrt(diag(Sigma.Beta.Gls))
list(RMSE_gls = RMSE.gls, SE_GLS = SE.GLS, Cov_Beta_GLS = Sigma.Beta.Gls)
```

### 2.3 Geostatistical prediction

#### 2.3.1 Ordinary Kriging (OK)

``` {r}
# Design a sparser prediction grid for demo
OK.GridPts <- spsample(Boundary.sp, n = 20000, type = "regular")
gridded(OK.GridPts) <- TRUE

# Use an allowable variogram for y
OK.Rain <- krige(y ~ 1, locations = rain.SpDF, newdata = OK.GridPts, model = vgmExp.y)

grid.arrange(
  spplot(OK.Rain, zcol = "var1.pred", main = "OK Predictions of Rainfall"),
  spplot(OK.Rain, zcol = "var1.var",  main = "OK Prediction Variance"),
  nrow = 1
)
```

#### 2.3.2 Regression Kriging (RK)

``` {r}
# Build newdata with covariates on grid
RK.GridPts <- spsample(Boundary.sp, n = 20000, type = "regular")
RK.GridPtsDF <- SpatialPointsDataFrame(
  RK.GridPts,
  data = data.frame(
    x1 = extract(DEM,  RK.GridPts, df = TRUE)[,2],
    x2 = extract(NDVI, RK.GridPts, df = TRUE)[,2]
  )
)
gridded(RK.GridPtsDF) <- TRUE

# One-step RK via gstat::krige, using a variogram for y | X
vgm.Emp.RK <- variogram(y ~ x1 + x2, locations = rain.SpDF, covariogram = FALSE)
vgmExp.RK  <- fit.variogram(vgm.Emp.RK, model = vgm(psill = 15000, model = "Exp", range = 60000, nugget = 500))

RK.Rain <- krige(y ~ x1 + x2, locations = rain.SpDF, newdata = RK.GridPtsDF, model = vgmExp.RK)

grid.arrange(
  spplot(RK.Rain, zcol = "var1.pred", main = "Regression Kriging — Prediction"),
  spplot(RK.Rain, zcol = "var1.var",  main = "Regression Kriging — Variance"),
  nrow = 1
)
```

### 2.4 Cross‑validation

We use leave‑one‑out CV to assess predictive performance and 5‑fold CV as a check.

``` {r}
# LOOCV
RK.Rain.CV <- krige.cv(y ~ x1 + x2, locations = rain.SpDF, model = vgmExp.RK)

ME  <- mean(RK.Rain.CV$residual)
MSE <- mean(RK.Rain.CV$residual^2)
RMSE <- sqrt(MSE)
R2   <- 1 - var(RK.Rain.CV$residual) / var(rain.SpDF$y)
c(ME = ME, RMSE = RMSE, R2 = R2)

# 5-fold CV
RK.Rain.5CV <- krige.cv(y ~ x1 + x2, locations = rain.SpDF, model = vgmExp.RK, nfold = 5)
c(ME_5fold = mean(RK.Rain.5CV$residual), RMSE_5fold = sqrt(mean(RK.Rain.5CV$residual^2)))
```

## 3) Assignment scaffold (fill and knit)

Below are minimal code stubs you can duplicate into separate sections to answer each task with metrics/tables/plots.

### Task 1 — Compare simple linear models: y \~ x1 vs y \~ x2

``` {r}
M1 <- lm(y ~ x1)
M2 <- lm(y ~ x2)

met <- function(m) {
  e  <- resid(m); n <- length(e); k <- length(coef(m)) - 1
  SSE <- sum(e^2); MSE <- SSE / (n - (k + 1)); RMSE <- sqrt(MSE)
  c(AIC = AIC(m), BIC = BIC(m), R2 = summary(m)$r.squared,
    AdjR2 = summary(m)$adj.r.squared, RMSE = RMSE)
}
rbind(M1 = met(M1), M2 = met(M2))
```

Explain which predictor performs better and why (use R²/Adj‑R², RMSE, residual diagnostics).

### Task 2 — Multiple regression: y \~ x1 + x2 vs Task 1 models

``` {r}
M3 <- LM1  # already fitted as y ~ x1 + x2
rbind(M1 = met(M1), M2 = met(M2), M3 = met(M3))
```

Discuss whether adding both predictors is justified (e.g., Adj‑R², RMSE, significance, multicollinearity signs).

### Task 3 — Compare OLS vs Geostatistical Regression (GLS)

``` {r}
RMSE_OLS <- RMSE
RMSE_GLS <- as.numeric(RMSE.gls)
c(RMSE_OLS = RMSE_OLS, RMSE_GLS = RMSE_GLS)
```

State which has better predictive power and interpret the practical impact.

### Task 4 — OK vs RK via LOOCV

``` {r}
OK.cv <- krige.cv(y ~ 1,        locations = rain.SpDF, model = vgmExp.y)
RK.cv <- krige.cv(y ~ x1 + x2,  locations = rain.SpDF, model = vgmExp.RK)

score <- function(cv) c(ME = mean(cv$residual), RMSE = sqrt(mean(cv$residual^2)),
                        R2 = 1 - var(cv$residual) / var(rain.SpDF$y))
rbind(OK = score(OK.cv), RK = score(RK.cv))
```

Interpret which method explains a higher proportion of variance and whether RK is worth the added complexity.

------------------------------------------------------------------------

## Appendix — Tips & Gotchas

-   Make sure CRS are consistent across points, rasters, and boundary.
-   Use sensible initial variogram parameters; refine if fits fail to converge.
-   If `spsample` appears slow, reduce grid size during prototyping.
-   For large rasters, consider `terra` package equivalents for speed.
-   When reporting, always pair metrics (e.g., RMSE) with diagnostic plots.
