---
title: "Linear and Geostatistical Regression (Swiss Rainfall)"
author: "David Reyes"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Overview

This R Markdown reproduces the **Linear and Geostatistical Regression** workflow based on the Swiss rainfall dataset described in the provided PDF tutorial.

It includes:
- OLS regression (`y ~ x1 + x2`)
- Geostatistical regression (GLS)
- Ordinary Kriging (OK)
- Regression Kriging (RK)
- Cross-validation metrics

---

# Load Packages

```{r packages}
library(sf)
library(raster)
library(sp)
library(spdep)
library(gridExtra)
library(cowplot)
library(gstat)
```

# Load and Prepare Data

```{r data}
# Set working directory
# setwd("~/Documents/Franky/Education/Foundation Course/Tutorial")

rain.DF <- read.table("sic_obs.dat", sep = ",", header = TRUE)
rain.SpDF <- rain.DF
coordinates(rain.SpDF) <- c("x_coor","y_coor")

Boundary.sf <- st_read("Swiss_Bound.shp", quiet = TRUE)
Boundary.sp <- as(Boundary.sf, "Spatial")

DEM  <- raster("DEM.grd")
NDVI <- raster("NDVI.grd")
```

# Extract Predictors

```{r predictors}
x1 <- extract(DEM, rain.SpDF, df = TRUE)[,2]
x2 <- extract(NDVI, rain.SpDF, df = TRUE)[,2]
y  <- rain.SpDF@data$rain_mm
X  <- cbind(1, x1, x2)

rain.SpDF@data$y <- y
rain.SpDF@data$x1 <- x1
rain.SpDF@data$x2 <- x2
```

# Fit OLS Model

```{r ols}
LM1 <- lm(y ~ x1 + x2)
summary(LM1)

y_cap <- predict(LM1)
e <- y - y_cap

SSE <- t(y - y_cap) %*% (y - y_cap)
K <- 2
MSE <- SSE / (length(y) - (K + 1))
RMSE <- sqrt(MSE); RMSE
```

# Variogram Models and GLS Iterations

```{r gls}
vgm.Emp.e <- variogram(e ~ 1, locations = rain.SpDF)
vgmSph.e <- fit.variogram(vgm.Emp.e, model = vgm(8000, "Sph", 60000, nugget = 500))

DMat <- spDists(rain.SpDF)
Sigma.ols1 <- variogramLine(vgmSph.e, dist_vector = DMat, covariance = TRUE)
Beta.gls1 <- solve(t(X) %*% solve(Sigma.ols1) %*% X) %*% (t(X) %*% solve(Sigma.ols1) %*% y)
e.gls1 <- y - (X %*% Beta.gls1)

B.gls.ite <- matrix(nrow = 100, ncol = 3)
for (i in 1:100) {
  vgm.Emp.gls1 <- variogram(e.gls1 ~ 1, locations = rain.SpDF)
  vgmExp.gls1 <- fit.variogram(vgm.Emp.gls1, model = vgm(8000, "Sph", 60000, nugget = 500))
  Sigma.gls1 <- variogramLine(vgmExp.gls1, dist_vector = DMat, covariance = TRUE)
  B.gls.ite[i, ] <- solve(t(X) %*% solve(Sigma.gls1) %*% X) %*% (t(X) %*% solve(Sigma.gls1) %*% y)
  e.gls1 <- y - (X %*% B.gls.ite[i, ])
}

B.gls <- B.gls.ite[100, ]  # Final GLS coefficients
e.gls <- y - (X %*% B.gls)
RMSE.gls <- sqrt(mean(e.gls^2)); RMSE.gls
```

# Ordinary Kriging and Regression Kriging

```{r kriging}
vgmExp.y <- fit.variogram(variogram(y ~ 1, locations = rain.SpDF), model = vgm(15000, "Exp", 60000, nugget = 500))
OK.GridPts <- spsample(Boundary.sp, n = 20000, type = "regular")
gridded(OK.GridPts) <- TRUE

OK.Rain <- krige(y ~ 1, locations = rain.SpDF, newdata = OK.GridPts, model = vgmExp.y)

RK.GridPtsDF <- SpatialPointsDataFrame(
  OK.GridPts,
  data.frame(x1 = extract(DEM, OK.GridPts, df = TRUE)[,2],
             x2 = extract(NDVI, OK.GridPts, df = TRUE)[,2])
)
gridded(RK.GridPtsDF) <- TRUE
RK.Rain <- krige(y ~ x1 + x2, locations = rain.SpDF, newdata = RK.GridPtsDF, model = vgmExp.y)
```

# Cross-validation

```{r cv}
RK.Rain.CV <- krige.cv(y ~ x1 + x2, locations = rain.SpDF, model = vgmExp.y)
ME  <- mean(RK.Rain.CV$residual)
MSE <- mean(RK.Rain.CV$residual^2)
RMSE <- sqrt(MSE)
R2 <- 1 - var(RK.Rain.CV$residual) / var(rain.SpDF$y)
data.frame(ME, MSE, RMSE, R2)
```

# Discussion

- **OLS regression** models the deterministic trend using DEM and NDVI.
- **Geostatistical regression (GLS)** improves estimation by accounting for spatial correlation in residuals.
- **Ordinary Kriging** assumes a constant mean across space.
- **Regression Kriging** combines both the deterministic trend and the stochastic spatial component for improved predictions.
- Cross-validation metrics (RMSE, R²) are used to evaluate predictive performance.
